#Severe Weather Events: Tornados most harmful while Floods have the greatest economic consequences

## Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

We studied the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database in order to find out which types of events are most harmful to the population health and which types of events have the greatest economic consequences. 

We hope that our findings will help you prepare for the server weather events in order to minimize their impact on the population and the economy.

## Data Processing
###Loading Necessary Libraries
```{r}
library(ggplot2)
library(dplyr)

```

###Downloading and Reading the data from the file 
```{r cache=TRUE}
destFile<-"repdata_data_StormData.csv.bz2"
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!file.exists(destFile)){
     download.file(fileURL ,destfile=destFile,method="auto")
} 
stormData <-read.csv("repdata_data_StormData.csv.bz2")
```

###Examining the data
```{r}
str(stormData)
head(stormData)
```

###Health Consequences
To evaluate the health impact, we calculate the total fatalities and total injuries of each event type. The variables EVTYPE, FATALITIES and INJURIES contain the information that we need.

A dataframe with only the needed columns
```{r cache=TRUE}
stormHealth <- stormData %>% select(EVTYPE, FATALITIES,INJURIES)
```

```{r  cache=TRUE}
fatalities <- stormHealth %>% select(EVTYPE, FATALITIES) %>% group_by(EVTYPE) %>% summarise(totalFatalities = sum(FATALITIES)) %>% arrange(-totalFatalities)

head(fatalities, 10)

injuries <- stormHealth %>% select(EVTYPE, INJURIES) %>% group_by(EVTYPE) %>% summarise(totalInjuries = sum(INJURIES)) %>% arrange(-totalInjuries)

head(injuries, 10)
```

###Economic Consequences
To evaluate the economic consequences we are interested in the following columns: EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP. The columns store the values of Property Damage and Crop damage and a multiplier for each value. 
The columns PROPDMGEXP and CROPDMGEXP contain the multiplier abbreviations. From the codebook, we can understand the following:

    Possible values: H,h,K,k,M,m,B,b,+,-,?,0,1,2,3,4,5,6,7,8, and blank character
    H,h = hundreds = 100
    K,k = kilos = thousands = 1,000
    M,m = millions = 1,000,000
    B,b = billions = 1,000,000,000
    (+) = 1
    (-) = 0
    (?) = 0
    blank/empty character = 0
    numeric 0..8 = 10


We construct a dataframe with only the needed columns
```{r}
stormDamage <- stormData %>% select(EVTYPE, PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP)
```

We add additional columns to our newly constructed data frame
- PROPMULT and CROPMULT columns will hold the values of the multipliers and will be used to calucate the dollar value of the damage
- To calculate the property damage dollar value PROPDMGUSD and crop damage dollar value CROPDMGUSD, we multiply the damage dollar number with the appropriate multiple for property and crop
- We then add dollar damage value in one field TOTALDMGUSD

```{r cache=TRUE}
stormDamage$PROPMULT = 0
stormDamage$CROPMULT = 0

stormDamage[stormDamage$PROPDMGEXP=="?",]$PROPMULT=0
stormDamage[stormDamage$PROPDMGEXP=="M",]$PROPMULT=1000000
stormDamage[stormDamage$PROPDMGEXP=="m",]$PROPMULT=1000000
stormDamage[stormDamage$PROPDMGEXP=="K",]$PROPMULT=1000
stormDamage[stormDamage$PROPDMGEXP=="H",]$PROPMULT=100
stormDamage[stormDamage$PROPDMGEXP=="h",]$PROPMULT=100
stormDamage[stormDamage$PROPDMGEXP=="B",]$PROPMULT=1000000000
stormDamage[stormDamage$PROPDMGEXP=="+",]$PROPMULT=0
stormDamage[stormDamage$PROPDMGEXP=="-",]$PROPMULT=0
stormDamage[stormDamage$PROPDMGEXP=="",]$PROPMULT=0

stormDamage$PROPDMGUSD = stormDamage$PROPDMG*stormDamage$PROPMULT

stormDamage[stormDamage$CROPDMGEXP=="?",]$CROPMULT=0
stormDamage[stormDamage$CROPDMGEXP=="M",]$CROPMULT=1000000
stormDamage[stormDamage$CROPDMGEXP=="m",]$CROPMULT=1000000
stormDamage[stormDamage$CROPDMGEXP=="K",]$CROPMULT=1000
stormDamage[stormDamage$CROPDMGEXP=="k",]$CROPMULT=1000
stormDamage[stormDamage$CROPDMGEXP=="B",]$CROPMULT=1000000000
stormDamage[stormDamage$CROPDMGEXP=="",]$CROPMULT=0

stormDamage$CROPDMGUSD = stormDamage$CROPDMG*stormDamage$CROPMULT

stormDamage$TOTALDMGUSD = stormDamage$CROPDMGUSD + stormDamage$PROPDMGUSD

```
We now group by event type and sort by damage dollar value descending 
```{r}
economical <- stormDamage %>% group_by(EVTYPE) %>% summarize(TOTALDMGEVENTTYPE = sum(TOTALDMGUSD))%>% arrange(-TOTALDMGEVENTTYPE) 

head(economical,10)
```

##Results

###Across the United States, which types of events are most harmful with respect to population health?
The top 10 events with the highest total fatalities are shown below
```{r}
fatalitiesPlot <- ggplot() + geom_bar(data = fatalities[1:10,], aes(x = EVTYPE, 
    y = totalFatalities, fill = interaction(totalFatalities, EVTYPE)), stat = "identity", 
    show.legend = F) + theme(axis.text.x = element_text(angle = 30, hjust = 1)) + 
    xlab("Harmful Events") + ylab("No. of fatalities") + ggtitle("Top 10 weather events causing fatalities") + 
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
print(fatalitiesPlot)
```
**We can see that Tornado is responsible of most fatalities.**

The top 10 events with the highest total injuries are shown below
```{r}
injuriesPlot <- ggplot() + geom_bar(data = injuries[1:10,], aes(x = EVTYPE, y = totalInjuries, 
    fill = interaction(totalInjuries, EVTYPE)), stat = "identity", show.legend = F) + 
    theme(axis.text.x = element_text(angle = 30, hjust = 1)) + xlab("Harmful Events") + 
    ylab("No. of Injuries") + ggtitle("Top 10 weather events causing Injuries") + 
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
print(injuriesPlot)
```

**We can see that Tornado is responsible of most injuries.**

###Across the United States, which types of events have the greatest economic consequences?
```{r}
economicalPlot<-ggplot() + geom_bar(data = economical[1:10,], aes(x = EVTYPE, y = TOTALDMGEVENTTYPE, fill = interaction(TOTALDMGEVENTTYPE, 
    EVTYPE)), stat = "identity", show.legend = F) + theme(axis.text.x = element_text(angle = 30, 
    hjust = 1)) + xlab("Event Type") + ylab("Total Damage Value in USD")
print(economicalPlot)
```

**We can see that Flood is responsible of highest economical impact.**

##Conclusion
Our study shows that Tornados are the most harmful and have the biggest number of fatalities and injuries while Floods have the greatest economic consequences on buildings and crops.