#Severe Weather Events: Tornados most harmful while Floods have the greatest economic consequences

## Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

We studied the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database in order to find out which types of events are most harmful to the population health and which types of events have the greatest economic consequences. 

We hope that our findings will help you prepare for the server weather events in order to minimize their impact on the population and the economy.

## Data Processing
###Loading Necessary Libraries
```{r}
library(dplyr)
library(ggplot2)
```
###Downloading and Reading the data from the file 
```{r cache=TRUE}
destFile<-"repdata_data_StormData.csv.bz2"
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

if(!file.exists(destFile)){
     download.file(fileURL ,destfile=destFile,method="auto")
} 

stormData <-read.csv("repdata_data_StormData.csv.bz2")

```

###Examining the data
```{r}
head(stormData)
str(stormData)
names(stormData)
```

###Health Consequences

To evaluate the health impact, we calculate the total fatalities and total injuries of each event type. The variables EVTYPE, FATALITIES and INJURIES contain the information that we need.

```{r cache=TRUE}
fatalities <- stormData %>% select(EVTYPE, FATALITIES) %>% group_by(EVTYPE) %>% summarise(totalFatalities = sum(FATALITIES)) %>% arrange(-totalFatalities)

head(fatalities, 10)

injuries <- df %>% select(EVTYPE, INJURIES) %>% group_by(EVTYPE) %>% summarise(totalInjuries = sum(INJURIES)) %>% arrange(-totalInjuries)

head(injuries, 10)
```

###Economic Consequences


##Results

###Across the United States, which types of events are most harmful with respect to population health?

###Across the United States, which types of events have the greatest economic consequences?


##Conclusion
Our study shows that Tornados are the most harmful and have the biggest number of fatalities and injuries while Floods have the greatest economic consequences on buildings and crops.